Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-23 prop_letter.tex
prop_letter.tex \section{Gleason} Updating...

When $\delta \ge \frac{1}{3}$, the situation abruptly
changes. Assuming a Hilbert space $\Hilb$ of dimension $3$, then for
any projector $P$ that projects onto an $n$-dimensional subspace of
$\Hilb$, it is straightforward to check that
$\bar{\mu}\left(P\right)=\left[n/3,n/3\right]$ is a valid
$\frac{1}{3}$-deterministic QIVPM. In other words, when the
uncertainty in measurements becomes so large, the Kochen-Specker
theorem no longer holds and it becomes possible to map every
observable to some (quite inaccurate) probability interval, and hence
the theory becomes non-contextual.

@@

While the finite-precision extension of the Kochen-Specker theorem
asserts the non-existence of a certain kind of QIVPMs, we also want
to understand what a QIVPM could be when it exists. A conventional
quantum probability measure~$\mu$ can easily be constructed using
the Born rule when one knows the current state of the system~\citep{Born1983,peres1995quantum,544199,Jaeger2007}.
Gleason's theorem asserts that the only state consistent with $\mu$
on all projectors is the one used to construct $\mu$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
In another word, the set of all quantum probability measures is one-to-one
corresponding to the set of states so that studying the set of states
is sufficient to understand all possible phenomenon regarding to quantum
probability. Although we can extend the Born rule to its finite precision
version easily, there are some subtleties extending Gleason's theorem.
In particular, there is a QIVPM such that no state can be consistent
with it on every projector. However, if we only focus on the Born
QIVPMs~$\bar{\mu}$, that is, QIVPM's induced by a state~$\rho$
according to the Born rule, we find that if we monotonically shrink
the length of intervals, the states consistent with $\bar{\mu}$ will
be closer and closer to $\rho$, and vise visa. As the result, although
a finite precision version of Gleason's theorem is stated, there might
be more rich phenomenon regarding to quantum probability when considering
finite precision measurement, and highlight the necessity to develop
a mathematical theory of quantum mechanics that intrinsically consistent
with the resources available for realistic measurement.

The first step toward a finite precision Born rule is to find a way
of replacing the computed probability by an interval. This can be
done by carefully picking a set of intervals and a rule to select
the correct interval when intervals are overlapping. For example,
consider the following collection of overlapping intervals: 
\begin{equation}
\begin{aligned}\mathscr{I}_{n}= & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\\
 & \cup\{\imposs,\necess\}
\end{aligned}
\end{equation}
which are overlapping intervals with length~$1/n$ except for the
intervals with length~$\left(2n\right)^{-1}$ at each end. To select
the correct interval from these overlapping intervals~$[\ell,r]$,
we define the non-overlapping interior part of a interval, $\textrm{Int}\left([\ell,r]\right)$,
as in Table~\ref{tab:I_n}.
\begin{table}
\caption{\label{tab:I_n}Intervals in $\mathscr{I}_{n}$ and their interior.
Although intervals~$[\ell,r]$ are overlapping, their interior~$\textrm{Int}\left([\ell,r]\right)$
are non-overlapping.}

\begin{tabular}{ccc}
\toprule 
\addlinespace
$k$ & $[\ell,r]$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
 & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0$ & $\left[0,\tfrac{1}{2n}\right]$ & $\left(0,\tfrac{0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
Any odd number & $\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
Any even number\tabularnewline
with $0<k<2n$\tabularnewline
\end{tabular} & \multirow{1}{*}{$\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$} & \multirow{1}{*}{$\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$}\tabularnewline\addlinespace
\midrule 
\addlinespace
$2n$ & $\left[\tfrac{2n-1}{2n},1\right]$ & $\left[\tfrac{2n-0.5}{2n},1\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
 & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
 To understand these intervals and their interior part more clear,
a special case $\mathscr{I}_{2}$ is also listed in Table~\ref{tab:I_2}.
\begin{table}
\caption{\label{tab:I_2}Intervals in $\mathscr{I}_{2}$ and their interior.
Although intervals~$[\ell,r]$ are overlapping, their interior~$\textrm{Int}\left([\ell,r]\right)$
are non-overlapping.}

@@

\noindent Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$,
it is easy to verify that
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
This implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, then we have
two possibilities:
\begin{itemize}
\item Their complements are orthogonal so that we can apply Eq.~(\ref{eq:iota-add})
on their complements.
\item One of them projects onto the subspace which the other one projects
onto.
\end{itemize}
\noindent No matter which case, Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
can be deduced easily. Therefore, $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is indeed a QIVPM. As a concrete example of Thm.~\ref{thm:Finite-precision-Born},
when $n=2$, two states~$\rho_{1}$ and $\rho_{0}$ can induce two
QIVPMs~$\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ and $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$,
where $\rho_{1}=\left(\proj{0}+\proj{1}+\proj{2}\right)/3$ and $\rho_{0}=\frac{7}{24}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{3}{8}\cdot\proj{2}$.
The interval-valued probability of the induced QIVPMs, $\iota_{2}\left(\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\right)$
and $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$,
for any projector $P$ in the minimal subspace of events containing
$\proj{0}$, $\proj{1}$, and $\proj{2}$, $\events_{C}$, is computed
straightforwardly in Table~\ref{tab:Compute-the-QIVPM}.
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$ based on $\mu_{\rho_{1}}^{\mathrm{B}}$
and $\mu_{\rho_{0}}^{\mathrm{B}}$ on $\events_{C}$.}

\begin{tabular}{ccccc}
\toprule 
\addlinespace
$P$ & $\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\right)$ & $\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $0$ & $\imposs$ & $0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{7}{24}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{2}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{3}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{2}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{1}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{0}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{17}{24}$ & $\left[\frac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $1$ & $\necess$ & $1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}

In the limiting case when the intervals are infinitely precise, i.e.,
$\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$
with its corresponding $\iota_{\infty}\left(x\right)=[x,x]$, we can
formulate Gleason's theorem in our framework as follow.

\begin{thm}[$\mathscr{I}_{\infty}$ Variant of Gleason]\label{cor:Gleason's}In
a Hilbert space $\Hilb$ of dimension $d\geq3$, given a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$,
the states consistent with it on every project, $\coreBorn\left(\bar{\mu},\events\right)$
has a unique element. \end{thm}

@@

\noindent For example, $\rho_{1}$ and $\rho_{0}$ are both totally
mixed states. Since $\left\Vert \rho_{1}-\rho_{0}\right\Vert \le\frac{1}{24}$,
we should have $\rho_{0}\in\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}},\events\right)$
and $\rho_{1}\in\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$,
and this is indeed the case for $P\in\events_{C}$ according to Table~\ref{tab:Compute-the-QIVPM}.
In general, as $n$ goes to infinity, only the state very close to
$\rho$ are guaranteed to belong to $\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}},\events\right)$,
and the states consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho$.

>>>>

When $\delta \ge \frac{1}{3}$, the situation abruptly
changes. Assuming a Hilbert space $\Hilb$ of dimension $3$, then for
any projector $P$ that projects onto an $n$-dimensional subspace of
$\Hilb$, it is straightforward to check that
$\bar{\mu}\left(P\right)=\left[\frac{n}{3},\frac{n}{3}\right]$ is a valid
$\frac{1}{3}$-deterministic QIVPM. In other words, when the
uncertainty in measurements becomes so large, the Kochen-Specker
theorem no longer holds and it becomes possible to map every
observable to some (quite inaccurate) probability interval, and hence
the theory becomes non-contextual.

@@

While the finite-precision extension of the Kochen-Specker theorem
asserts the non-existence of a certain kind of QIVPMs, we also want
to understand what a QIVPM could be when it exists. A conventional
quantum probability measure~$\mu$ can easily be constructed using
the Born rule when one knows the current state of the system~\citep{Born1983,peres1995quantum,544199,Jaeger2007}.
Gleason's theorem asserts that the only state consistent with $\mu$
on all projectors is the one used to construct $\mu$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
In another word, the set of all quantum probability measures is one-to-one
corresponding to the set of states so that studying the set of states
is sufficient to understand all possible phenomenon regarding to quantum
probability. Although we can extend the Born rule to its finite precision
version easily, there are some subtleties extending Gleason's theorem.
In particular, there is a QIVPM such that no state can be consistent
with it on every projector. However, if we only focus on the Born
QIVPMs~$\bar{\mu}$, that is, QIVPM's induced by a state~$\rho$
according to the Born rule, we find that if we monotonically shrink
the length of intervals, the states consistent with $\bar{\mu}$ will
be closer and closer to $\rho$, and vise visa. In another word, although
we can prove a finite precision version of Gleason's theorem, there
might be more rich phenomenon regarding to quantum probability when
considering finite precision measurement. Our result also highlight
the necessity to develop a mathematical theory of quantum mechanics
that intrinsically consistent with the resources available for realistic
measurement.

The first step toward a finite precision Born rule is to find a way
of replacing the computed probability by an interval. This can be
done by carefully picking a set of intervals and a rule to select
the correct interval when intervals are overlapping. For example,
consider the following collection of overlapping intervals: 
\begin{equation}
\begin{aligned}\mathscr{I}_{n}= & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\\
 & \cup\{\imposs,\necess\}
\end{aligned}
\end{equation}
which are overlapping intervals with length~$1/n$ except for the
intervals with length~$\left(2n\right)^{-1}$ at each end. As in
Table~\ref{tab:I_n}, To select the correct interval from these overlapping
intervals~$[\ell,r]$, we define the non-overlapping interior part
of a interval, $\textrm{Int}\left([\ell,r]\right)$.
\begin{table}
\caption{\label{tab:I_n}Intervals in $\mathscr{I}_{n}$ and their interior.
Although intervals~$[\ell,r]$ are overlapping, their interior~$\textrm{Int}\left([\ell,r]\right)$
are non-overlapping. See Table~\ref{tab:I_2} for a concrete example
when $n=2$.}

\begin{tabular}{ccc}
\toprule 
\addlinespace
$k$ & $[\ell,r]$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
 & $\imposs$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0$ & $\left[0,\tfrac{1}{2n}\right]$ & $\left(0,\tfrac{0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
Any odd number & $\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
Any even number\tabularnewline
with $0<k<2n$\tabularnewline
\end{tabular} & \multirow{1}{*}{$\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$} & \multirow{1}{*}{$\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$}\tabularnewline\addlinespace
\midrule 
\addlinespace
$2n$ & $\left[\tfrac{2n-1}{2n},1\right]$ & $\left[\tfrac{2n-0.5}{2n},1\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
 & $\necess$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
 Table~\ref{tab:I_2} lists a special case $\mathscr{I}_{2}$ to
illustrate these intervals and their interior parts more clearly.
\begin{table}
\caption{\label{tab:I_2}Intervals in $\mathscr{I}_{2}$ and their interior.
Although intervals~$[\ell,r]$ are overlapping, their interior~$\textrm{Int}\left([\ell,r]\right)$
are non-overlapping.}

@@

\noindent Since $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$ clearly satisfies
Eqs.(\ref{eq:QIVPM-constraints}), the condition really need to be
checked is Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$, it
is easy to verify that
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
This implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements are orthogonal so that we can apply Eq.~(\ref{eq:iota-add})
on their complements.
\item One of them projects onto the subspace which the other one projects
onto.
\end{itemize}
\noindent No matter which case, Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
can be deduced easily. Therefore, $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is indeed a QIVPM.

As a concrete example of Thm.~\ref{thm:Finite-precision-Born}, when
$n=2$, two states~$\rho_{0}=\left(\proj{0}+\proj{1}+\proj{2}\right)/3$
and $\rho_{1}=\frac{7}{24}\cdot\proj{0}+\frac{1}{3}\cdot\proj{1}+\frac{3}{8}\cdot\proj{2}$
can induce two QIVPMs~$\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$. Given any projector
$P$ in the minimal subspace of events containing $\proj{0}$, $\proj{1}$,
and $\proj{2}$, $\events_{C}$, the interval-valued probability of
the induced QIVPMs, $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$
and $\iota_{2}\left(\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\right)$,
is computed straightforwardly in Table~\ref{tab:Compute-the-QIVPM}.
\begin{table}
\caption{\label{tab:Compute-the-QIVPM}Compute $\iota_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}}$
and $\iota_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}}$ based on $\mu_{\rho_{0}}^{\mathrm{B}}$
and $\mu_{\rho_{1}}^{\mathrm{B}}$ on $\events_{C}$.}

\begin{tabular}{ccccc}
\toprule 
\addlinespace
$P$ & $\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\rho_{0}}^{\mathrm{B}}\left(P\right)\right)$ & $\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)$ & $\iota_{2}\left(\mu_{\rho_{1}}^{\mathrm{B}}\left(P\right)\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
$\mathbb{0}$ & $0$ & $\imposs$ & $0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{0}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{7}{24}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{1}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\proj{2}$ & $\frac{1}{3}$ & $\left[0,\frac{1}{2}\right]$ & $\frac{3}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{2}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{5}{8}$ & $\left[\tfrac{1}{4},\tfrac{3}{4}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{1}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}-\proj{0}$ & $\frac{2}{3}$ & $\left[\frac{1}{2},1\right]$ & $\frac{17}{24}$ & $\left[\frac{1}{2},1\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$\mathbb{1}$ & $1$ & $\necess$ & $1$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}

In the limiting case when the intervals are infinitely precise, i.e.,
$\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
we can formulate Gleason's theorem in our framework as follow.

\begin{thm}[$\mathscr{I}_{\infty}$ Variant of Gleason]\label{cor:Gleason's}In
a Hilbert space $\Hilb$ of dimension $d\geq3$, given a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$,
the states consistent with it on every project, $\coreBorn\left(\bar{\mu},\events\right)$,
has a unique element. \end{thm}

@@

\noindent For example, $\rho_{0}$ and $\rho_{1}$ are both totally
mixed states. Since $\left\Vert \rho_{0}-\rho_{1}\right\Vert \le\frac{1}{24}$,
we should have $\rho_{1}\in\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$
and $\rho_{0}\in\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{1}}^{\mathrm{B}},\events\right)$,
and this is indeed the case for $P\in\events_{C}$ according to Table~\ref{tab:Compute-the-QIVPM}.
In general, as $n$ goes to infinity, only the state very close to
$\rho$ are guaranteed to belong to $\coreBorn\left(\widehat{\iota}_{2}\circ\mu_{\rho_{0}}^{\mathrm{B}},\events\right)$,
and the states consistent with $\iota_{n}\circ\mu_{\rho}^{\mathrm{B}}$
will become closer and closer to $\rho$.
